---
title: "California Housing Project - Group 5"
author: Hilda Ndambuki(053837), Jakes Chege(222082), Joseph Kosen(133198), Joshua
  Ngumi(136346), Linet Wanjira(193883)
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

# Introduction

## Background

Housing markets are among the most economically significant and data-rich 
domains in applied statistics. The price of a residential property is 
determined by a complex web of factors — location, neighbourhood income 
levels, structural characteristics, and population density — whose 
relationships with price are rarely linear and rarely independent of one 
another. A district's median income, for instance, does not have a fixed 
additive effect on house value; its influence is amplified or dampened 
depending on where that district sits geographically. Capturing these 
nonlinear, spatially-varying dynamics is precisely where classical linear 
models fall short and where flexible statistical and machine learning 
approaches become essential.

## Dataset

This project uses the **California Housing dataset**, derived from the 1990 
U.S. Census. Each observation represents a census block group — the smallest 
geographical unit for which the Census Bureau publishes sample data — 
containing between 600 and 3,000 people. The dataset comprises 20,640 
observations and 9 variables:

| Variable | Description |
|---|---|
| `longitude` / `latitude` | Geographic coordinates of the block group centroid |
| `housing_median_age` | Median age of houses within the block group |
| `total_rooms` | Total number of rooms across all households |
| `total_bedrooms` | Total number of bedrooms across all households |
| `population` | Total population of the block group |
| `households` | Number of households |
| `median_income` | Median household income (in tens of thousands USD) |
| `median_house_value` | **Target** — median house value (USD) |

The raw count variables (`total_rooms`, `total_bedrooms`, `population`) are 
meaningless in isolation without knowing how many households share them. 
We therefore engineer per-household ratios — `rooms_per_household`, 
`bedrooms_per_room`, and `population_per_household` — which carry genuine 
economic meaning and replace their parent columns in all models.

## Research Questions

This project is structured around three guided research questions that 
together span the full spectrum from predictive accuracy to interpretable 
nonlinear inference:

1. **Can we predict house prices more accurately using an ensemble of 
nonlinear models?** We build and evaluate Random Forest, XGBoost, and a 
Neural Network individually, then combine them into an averaging ensemble 
to test whether model combination yields further gains over any single learner.

2. **What are the most important nonlinear interactions between location and 
income?** Using Generalised Additive Models with tensor product smooths and 
Random Forest partial dependence plots, we investigate whether and how the 
relationship between median income and house value varies across California's 
geography.

3. **Which model family — parametric regression, generalised nonparametric 
regression (GNPR), or machine learning — provides the best predictive 
accuracy?** We conduct a structured comparison across all three families, 
evaluating each on RMSE, MAE, and R² on a common held-out test set with a 
common feature set, so that differences in performance reflect model class 
rather than information asymmetry.

## Analytical Approach

Models are grouped into three families that reflect a deliberate 
bias–variance–interpretability trade-off:

| Family | Models | Key Property |
|---|---|---|
| Parametric | Linear Regression, Polynomial Regression | Strong assumptions, high interpretability |
| GNPR | Step Functions, Natural Splines, GAM, LOESS, Kernel | Flexible smooth functions, retains interpretability |
| Machine Learning | Random Forest, XGBoost, Neural Net, Ensemble | Maximum flexibility, lower interpretability |

All models are trained on an identical seven-feature set and evaluated on 
the same 20% held-out test set. This design ensures the final comparison 
table answers Question 3 cleanly — any difference in RMSE reflects what the 
model can do, not what information it was given.


# Setup: Install & Load Packages

Each package serves a specific role:

- **tidyverse / ggplot2** — data wrangling and visualisation
- **caret** — unified train/test split and pre-processing
- **randomForest** — Random Forest ensemble
- **xgboost** — gradient-boosted trees
- **nnet** — single hidden-layer neural network
- **mgcv** — Generalised Additive Models (GAM)
- **splines** — natural cubic splines inside `lm()`
- **corrplot** — correlation heatmap
- **pdp / vip** — partial-dependence plots and variable importance
- **maps** — California county outlines for geographic plots

```{r packages}
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "nnet",
          "mgcv", "splines", "corrplot", "ggplot2", "dplyr",
          "pdp", "vip", "maps", "boot")

for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
  library(p, character.only = TRUE)
}
```

---

# Load Data

Reads the raw `.data` file (no header) and attaches descriptive column names so every subsequent step is readable.

```{r load-data}
data <- read.csv(
  "C:/Users/ADMIN/Downloads/cal_housing.data",
  header = FALSE
)

colnames(data) <- c("longitude", "latitude", "housing_median_age",
                    "total_rooms", "total_bedrooms", "population",
                    "households", "median_income", "median_house_value")

dim(data)
head(data)
```

# Exploratory Data Analysis (EDA)

## Summary Statistics

Gives a quick numerical overview — min, max, median, and quartiles for every variable. Helps spot skewed distributions or impossible values before modelling.

```{r eda-summary}
summary(data)
```

## Distribution of Median House Value 

The response variable is right-skewed, with a hard cap visible at \$500,000 (a data-collection artefact). Nonparametric and nonlinear models handle this naturally without any transformation.

```{r eda-target-hist}
ggplot(data, aes(x = median_house_value)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of Median House Value",
       x = "Median House Value (USD)", y = "Count") +
  theme_minimal()
```

## Distribution of Median Income

Income is right-skewed. GAMs, splines, and tree models will learn the nonlinear income–price relationship directly from the data — no log transformation is required.

```{r eda-income-hist}
ggplot(data, aes(x = median_income)) +
  geom_histogram(bins = 50, fill = "darkorange", colour = "white") +
  labs(title = "Distribution of Median Income",
       x = "Median Income (tens of thousands USD)", y = "Count") +
  theme_minimal()
```

## Boxplots of All Raw Variables

One boxplot per variable reveals spread and outliers. Long upper whiskers on room and population counts signal heavy right skew — these will be addressed via feature engineering rather than transformation.

```{r eda-boxplots, fig.height=8}
data %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "tomato", outlier.alpha = 0.2, outlier.size = 0.5) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(title = "Boxplots of All Variables") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## Correlation Matrix

Shows which pairs of variables move together. `median_income` has the strongest positive correlation with house value. `latitude` and `longitude` show moderate correlations, hinting at spatial structure.

```{r eda-correlation}
cor_matrix <- cor(data, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.75, addCoef.col = "black", number.cex = 0.6,
         title = "Correlation Matrix", mar = c(0, 0, 1, 0))
```

## Geographic Distribution of House Values

Every district plotted by longitude/latitude and coloured by house value. The coastal premium is immediately visible — the most expensive properties cluster near San Francisco and Los Angeles.

```{r eda-map}
ca_counties <- map_data("county") |> subset(region == "california")

ggplot(data, aes(x = longitude, y = latitude)) +
  geom_polygon(data = ca_counties, aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray70") +
  geom_point(aes(colour = median_house_value), alpha = 0.3, size = 0.4) +
  coord_fixed(1.3) +
  scale_colour_viridis_c(name = "House Value ($)") +
  labs(title = "California House Values by Location",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

# Data Cleaning

## Missing Values

`total_bedrooms` has 207 missing entries. We impute with the column median — robust to the skewed distribution.

```{r clean-missing}
cat("Missing values per column:\n")
print(colSums(is.na(data)))

for (col in colnames(data)) {
  if (any(is.na(data[[col]]))) {
    data[[col]][is.na(data[[col]])] <- median(data[[col]], na.rm = TRUE)
  }
}
cat("After imputation:", sum(is.na(data)), "NAs remaining\n")
```

## Duplicates

```{r clean-duplicates}
cat("Duplicate rows found:", sum(duplicated(data)), "\n")
data <- data[!duplicated(data), ]
```

## Outlier Capping

Caps the top and bottom 1% of the raw count columns. These outliers can distort tree splits. Note: we cap *before* feature engineering so the derived ratios also benefit.

```{r clean-outliers}
cap_outliers <- function(x) {
  lo <- quantile(x, 0.01, na.rm = TRUE)
  hi <- quantile(x, 0.99, na.rm = TRUE)
  pmax(pmin(x, hi), lo)
}

skewed_cols <- c("total_rooms", "total_bedrooms", "population", "households")
data[, skewed_cols] <- lapply(data[, skewed_cols], cap_outliers)
```

---

# Preprocessing

## Feature Engineering

Raw totals (rooms, bedrooms, population) are economically meaningless without knowing how many households share them. We replace each raw count with its per-household ratio. The original raw columns are **dropped from all models** — the engineered features supersede them entirely.

```{r preproc-engineering}
data$rooms_per_household      <- data$total_rooms    / data$households
data$bedrooms_per_room        <- data$total_bedrooms / data$total_rooms
data$population_per_household <- data$population     / data$households
```

## Define the Canonical Feature Set

A single feature set is used across **all models** — parametric, GNPR, and machine learning alike. This ensures that any difference in performance reflects model class, not information asymmetry.

The raw parent columns (`total_rooms`, `total_bedrooms`, `population`, `households`) are excluded because they are fully represented by the engineered ratios above. `income_category` is retained for EDA visualisation only, not for modelling.

```{r preproc-features}
MODEL_FEATURES <- c(
  "longitude", "latitude", "housing_median_age",
  "median_income",
  "rooms_per_household",
  "bedrooms_per_room",
  "population_per_household"
)

cat("Features used in all models:\n")
print(MODEL_FEATURES)
```

## Income Category (EDA Only)

Used only to colour the geographic facet plot below — not included in any model.

```{r preproc-income-cat}
data$income_category <- cut(
  data$median_income,
  breaks = c(-Inf, 2.5, 4.5, Inf),
  labels = c("Low", "Medium", "High")
)
table(data$income_category)
```

```{r eda-map-income, fig.height=5}
ggplot(data, aes(x = longitude, y = latitude)) +
  geom_polygon(data = ca_counties, aes(x = long, y = lat, group = group),
               fill = "gray95", color = "gray70") +
  geom_point(aes(colour = median_house_value), alpha = 0.3, size = 0.4) +
  coord_fixed(1.3) +
  scale_colour_viridis_c(name = "House Value ($)") +
  facet_wrap(~income_category) +
  labs(title = "House Values by Location, Faceted by Income Category",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

## Train / Test Split (80 / 20)

```{r preproc-split}
set.seed(42)
train_idx <- createDataPartition(data$median_house_value, p = 0.8, list = FALSE)

# Formula-based models (LM, GAM, splines, LOESS) use train_orig / test_orig
train_orig <- data[ train_idx, ]
test_orig  <- data[-train_idx, ]

cat("Train rows:", nrow(train_orig), "| Test rows:", nrow(test_orig), "\n")
```

## Scaling for ML Models (RF, XGBoost, Neural Net)

Tree models do not need scaling, but the neural network does. We centre and scale only the model features — **not** the raw parent columns.

The response `median_house_value` is kept on its original dollar scale for all models so RMSE and MAE are directly comparable.

```{r preproc-scale}
# Subset to only the chosen features + target
train_ml <- train_orig[, c(MODEL_FEATURES, "median_house_value")]
test_ml  <- test_orig[,  c(MODEL_FEATURES, "median_house_value")]

# Centre and scale features using training statistics only
preproc <- preProcess(train_ml[, MODEL_FEATURES], method = c("center", "scale"))

train_s <- predict(preproc, train_ml[, MODEL_FEATURES])
test_s  <- predict(preproc, test_ml[,  MODEL_FEATURES])

# Attach the unscaled response — all eval_model() calls use original dollar values
train_s$median_house_value <- train_ml$median_house_value
test_s$median_house_value  <- test_ml$median_house_value

# Matrix form for RF and XGBoost
X_train <- as.matrix(train_s[, MODEL_FEATURES])
y_train <- train_s$median_house_value
X_test  <- as.matrix(test_s[,  MODEL_FEATURES])
y_test  <- test_s$median_house_value

# Confirm y_test matches test_orig target — scales must be identical
stopifnot(all.equal(y_test, test_orig$median_house_value))
cat("Scale check passed: y_test matches test_orig$median_house_value\n")
```

---

# Helper: Evaluation Metrics

A single function computes **RMSE**, **MAE**, and **R²** on the original dollar scale. Using this for every model guarantees a fair, like-for-like comparison table.

```{r helper-metrics}
eval_model <- function(actual, predicted, model_name) {
  rmse_val <- sqrt(mean((actual - predicted)^2))
  mae_val  <- mean(abs(actual - predicted))
  r2_val   <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  cat(sprintf("\n--- %-40s RMSE: %8.0f | MAE: %8.0f | R2: %.4f\n",
              paste0(model_name, " ---"), rmse_val, mae_val, r2_val))
  data.frame(Model    = model_name,
             RMSE     = round(rmse_val),
             MAE      = round(mae_val),
             R2       = round(r2_val, 4))
}

all_results <- list()
```

---

# Question 1 — Can Nonlinear Ensembles Predict House Prices More Accurately?

We first establish a **parametric baseline** (linear regression), then build three nonlinear ML models, and finally combine them into an **averaging ensemble**. The ensemble result directly answers Research Question 1.

## Baseline — Linear Regression

The simplest benchmark. Assumes each predictor has a straight-line relationship with house value. Expected to underfit because housing prices are known to be nonlinear.

```{r lm-baseline}
lm_model <- lm(
  median_house_value ~ .,
  data = train_orig[, c(MODEL_FEATURES, "median_house_value")]
)

lm_preds <- predict(lm_model, newdata = test_orig)
all_results$lm <- eval_model(test_orig$median_house_value, lm_preds,
                             "Linear Regression (baseline)")
summary(lm_model)
```

## Polynomial Regression
### Selecting the best degree of polynomial through Cross Validating
```{r}
set.seed(42)
cv_errors <- sapply(1:6, function(d) {
  fit <- glm(median_house_value ~ poly(median_income, d), data = train_orig)
  cv.glm(train_orig, fit, K = 10)$delta[1]
})

plot(1:6, cv_errors, type = "b", pch = 19,
     xlab = "Degree", ylab = "10-Fold CV Error",
     main = "Polynomial Degree Selection via CV")
best_d <- which.min(cv_errors)
cat("Best degree:", best_d, "\n")
```

Fitting a 6-degree polynomial of median income on the median house value
```{r}
# Using poly() 
poly6 <- lm(median_house_value ~ poly(median_income, 6), data = train_orig)
summary(poly6)
```

Fitting a polynomial of degree 6 on income and degree 3 on other predictors
```{r}
poly_model1 <- lm(
  median_house_value ~
  poly(median_income, 6, raw = FALSE) +
  poly(latitude,  3, raw = FALSE) +
  poly(longitude, 3, raw = FALSE) +
  poly(housing_median_age,  3, raw = FALSE) +
  poly(bedrooms_per_room, 3, raw = FALSE) +
  poly(population_per_household, 3, raw = FALSE),
  data = train_orig
)

poly_preds1  <- predict(poly_model1, newdata = test_orig)
all_results$poly  <- eval_model(test_orig$median_house_value, poly_preds1,
      "Polynomial Regression (degree per CV)")

summary(poly_model1)
```
Visualizing how degree six on income fits the data
```{r}
age_grid <- data.frame(median_income = seq(min(train_orig$median_income),
                                            max(train_orig$median_income),
                                            length.out = 300))
poly_pred2 <- predict(poly6, newdata = age_grid, se = TRUE)

# 95% CI (as shown in the lecture notes)
age_grid$fit   <- poly_pred2$fit
age_grid$upper <- poly_pred2$fit + 1.96 * poly_pred2$se.fit
age_grid$lower <- poly_pred2$fit - 1.96 * poly_pred2$se.fit

ggplot(train_orig, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.1, colour = "gray50") +
  geom_line(data = age_grid, aes(y = fit), colour = "blue", linewidth = 1) +
  geom_ribbon(data = age_grid, aes(y = fit, ymin = lower, ymax = upper),
              alpha = 0.2, fill = "blue") +
  labs(title = "Degree-6 Polynomial Regression: Income vs House Value",
       x = "Median Income", y = "Median House Value") +
  theme_minimal()
```


Comparing between degree 4 and 6 of polynomials of median income on median house value.
```{r}
poly4 <- lm(median_house_value ~ poly(median_income, 4), data = train_orig)
poly6 <- lm(median_house_value ~ poly(median_income, 6), data = train_orig)

all_results$poly4 <- eval_model(test_orig$median_house_value,
                                predict(poly4, test_orig), "Polynomial (deg-4)")
all_results$poly6 <- eval_model(test_orig$median_house_value,
                                predict(poly6, test_orig), "Polynomial (deg-6)")
```

#### Checking the Bias Variance Trade-off

```{r}
# Collect train and test errors for each degree
bv_results <- lapply(1:6, function(d) {
  fit        <- lm(median_house_value ~ poly(median_income, d), data = train_orig)
  train_rmse <- sqrt(mean((train_orig$median_house_value - predict(fit, train_orig))^2))
  test_rmse  <- sqrt(mean((test_orig$median_house_value  - predict(fit, test_orig))^2))
  data.frame(Degree = d, Train_RMSE = train_rmse, Test_RMSE = test_rmse)
})

bv_df <- do.call(rbind, bv_results)

bv_df %>%
  pivot_longer(cols = c(Train_RMSE, Test_RMSE),
               names_to = "Set", values_to = "RMSE") %>%
  ggplot(aes(x = Degree, y = RMSE, colour = Set)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:6) +
  labs(title = "Bias-Variance Trade-off: Polynomial Degree vs RMSE",
       x = "Polynomial Degree", y = "RMSE") +
  theme_minimal()
```
MSE = Variance + Bias^2
RMSE = squareroot(mse)
We can still see that the least RMSE is at degree 6
The test rmse scores are higher because they adopted the biases from the train selection

## Random Forest

Builds 300 decision trees on random subsets of data and features, then averages their predictions. The ensemble is more accurate and robust than any single tree.

```{r rf-model}
set.seed(42)
rf_model <- randomForest(
  x          = X_train,
  y          = y_train,
  ntree      = 200,
  mtry       = floor(sqrt(ncol(X_train))),
  importance = TRUE
)

rf_preds <- predict(rf_model, X_test)
all_results$rf <- eval_model(y_test, rf_preds, "Random Forest")
```

### Random Forest — Feature Importance

`%IncMSE`: how much worse the model gets when each feature is randomly shuffled. Higher values mean the feature is more important.

```{r rf-importance, fig.height=5}
imp_df <- as.data.frame(importance(rf_model, type = 1))
imp_df$Feature <- rownames(imp_df)
imp_df <- imp_df[is.finite(imp_df[, 1]), ]
colnames(imp_df)[1] <- "IncMSE"

imp_df %>%
  slice_max(IncMSE, n = 10) %>%
  mutate(Feature = reorder(Feature, IncMSE)) %>%
  ggplot(aes(x = Feature, y = IncMSE, fill = IncMSE)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_viridis_c() +
  labs(title = "Random Forest: Feature Importance (%IncMSE)",
       x = NULL, y = "% Increase in MSE when shuffled") +
  theme_minimal() +
  theme(legend.position = "none")
```


## XGBoost

Gradient boosting builds trees sequentially, each correcting the errors of the previous one. Often the strongest off-the-shelf model for tabular regression tasks.

```{r xgb-model}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.05,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(42)
xgb_model <- xgb.train(
  params    = xgb_params,
  data      = dtrain,
  nrounds   = 300,
  watchlist = list(train = dtrain, test = dtest),
  verbose   = 0
)

xgb_preds <- predict(xgb_model, dtest)
all_results$xgb <- eval_model(y_test, xgb_preds, "XGBoost")
```


### XGBoost — Feature Importance

```{r xgb-importance, fig.height=5}
xgb_imp <- xgb.importance(model = xgb_model)
xgb.plot.importance(xgb_imp, top_n = 10,
                    main = "XGBoost: Feature Importance (Gain)")
```

## Neural Network
```{r nn-model}
set.seed(42)
nn_model <- nnet(
  median_house_value ~ .,
  data   = train_s,
  size   = 10,
  linout = TRUE,
  maxit  = 500,
  decay  = 0.01
)

nn_preds <- as.vector(predict(nn_model, test_s[, MODEL_FEATURES]))
all_results$nn <- eval_model(y_test, nn_preds, "Neural Network (nnet)")
```

## Ensemble: Average of RF + XGBoost 

Combines the nonlinear models by simple averaging. If the models make different types of errors, averaging smooths them out and produces a more stable prediction. This directly answers Research Question 1.

```{r ensemble-model}
ensemble_preds <- (rf_preds + xgb_preds) / 2
all_results$ensemble <- eval_model(y_test, ensemble_preds,
                                   "Ensemble (RF + XGB  avg)")
```

# Question 2 — Nonlinear Interactions Between Location and Income

This section investigates *how* income and geography jointly determine house prices — the mechanism behind the predictions, not just their accuracy.

## Partial Dependence: Income × Latitude (Random Forest)

A 2-D partial dependence plot marginalises over all other variables and shows how the predicted price changes as income and latitude vary simultaneously. If the surface is not flat, the interaction is genuinely nonlinear

```{r}
# Fast 2-D interaction plot: RF predictions on a grid
# Build a grid of median_income x latitude values
income_grid <- quantile(train_orig$median_income, probs = seq(0.05, 0.95, length.out = 20))
lat_grid    <- quantile(train_orig$latitude,      probs = seq(0.05, 0.95, length.out = 20))

grid_df <- expand.grid(
  median_income = income_grid,
  latitude      = lat_grid
)

# Fix all other features at their median
for (f in setdiff(MODEL_FEATURES, c("median_income", "latitude"))) {
  grid_df[[f]] <- median(train_orig[[f]], na.rm = TRUE)
}

# Scale the grid the same way training data was scaled
grid_scaled  <- predict(preproc, grid_df[, MODEL_FEATURES])
grid_preds   <- predict(rf_model, as.matrix(grid_scaled))
grid_df$pred <- grid_preds

# Plot
ggplot(grid_df, aes(x = median_income, y = latitude, fill = pred)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted\nHouse Value ($)", direction = -1) +
  labs(
    title = "RF Predicted House Value: Income × Latitude Interaction",
    x     = "Median Income (tens of thousands USD)",
    y     = "Latitude"
  ) +
  theme_minimal()

```

> **Interpretation:** The steepest price gradient (brightest region) occurs where *both* income and latitude are high — coastal Northern California. A unit increase in income buys more house value in high-latitude coastal areas than in inland Southern California. This is a genuine nonlinear interaction invisible to additive models.

## GAM Interaction Surface: Income × Latitude

The GAM `ti()` tensor smooth estimates the interaction parametrically. `vis.gam()` renders the fitted surface as a contour plot. Contour lines that are not parallel and evenly spaced confirm a non-additive relationship.

```{r gam-interaction-surface, fig.height=6}
# Fit a focused interaction GAM for visualisation
gam_interact_vis <- gam(
  median_house_value ~
    s(median_income, k = 10) +
    s(latitude,      k = 10) +
    ti(median_income, latitude, k = 6),
  data   = train_orig[, c(MODEL_FEATURES, "median_house_value")],
  method = "REML"
)

vis.gam(
  gam_interact_vis,
  view      = c("median_income", "latitude"),
  plot.type = "contour",
  color     = "terrain",
  n.grid    = 40,
  main      = "GAM Interaction Surface: Income × Latitude",
  xlab      = "Median Income", ylab = "Latitude"
)
```

> **Interpretation:** The contour lines tilt and compress in the upper-right quadrant (high income, high latitude), confirming the same pattern found in the PDP: the income premium is location-dependent, not a fixed additive effect.



# Question 3 — GNPR vs. Parametric vs. Machine Learning


## GNPR - Step Functions

```{r}
# Cut median_income into bins — step function approach
train_step <- train_orig[, c(MODEL_FEATURES, "median_house_value")]
test_step  <- test_orig[,  c(MODEL_FEATURES, "median_house_value")]

# Create step/bin cuts on training data, apply same breaks to test
train_step$income_step <- cut(train_orig$median_income,
                               breaks = quantile(train_orig$median_income,
                                                 probs = seq(0, 1, by = 0.25)),
                               include.lowest = TRUE)

test_step$income_step  <- cut(test_orig$median_income,
                               breaks = quantile(train_orig$median_income,
                                                 probs = seq(0, 1, by = 0.25)),
                               include.lowest = TRUE)

step_model <- lm(
  median_house_value ~
    income_step +
    latitude + longitude + housing_median_age +
    rooms_per_household + bedrooms_per_room + population_per_household,
  data = train_step
)

step_preds <- predict(step_model, newdata = test_step)
all_results$step <- eval_model(test_orig$median_house_value, step_preds,
                               "Step Function (income bins)")
summary(step_model)

```
## GNPR - Smoothing Splines
Smoothing splines only work on a single predictor. The standard approach for a multivariate setting is to fit one per key predictor and compare, or use it as a component inside a GAM.

```{r}
# Smoothing spline on median_income only — CV selects lambda automatically
smooth_sp <- smooth.spline(
  x = train_orig$median_income,
  y = train_orig$median_house_value,
  cv = TRUE   # leave-one-out CV to choose lambda
)

cat("Smoothing spline effective df:", smooth_sp$df, "\n")

# Predict and plot
ss_preds_train <- predict(smooth_sp, x = train_orig$median_income)$y

ggplot(train_orig, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.1, size = 0.4, colour = "steelblue") +
  geom_line(aes(y = ss_preds_train), colour = "red", linewidth = 1) +
  labs(title = "Smoothing Spline: Median Income vs House Value",
       subtitle = paste("CV-selected effective df =", round(smooth_sp$df, 2)),
       x = "Median Income", y = "Median House Value ($)") +
  theme_minimal()
```


## GNPR — Natural Cubic Splines in Linear Regression

Instead of the full GAM machinery, we expand each predictor into a set of spline basis functions inside a plain `lm()`. This is faster and still captures curvature, but it does not automatically select the degree of smoothness.
The degrees of freedom were hard coded and arbitrarily chosen

```{r spline-model}
spline_model <- lm(
  median_house_value ~
    ns(median_income,            df = 5) +
    ns(latitude,                 df = 5) +
    ns(longitude,                df = 5) +
    ns(housing_median_age,       df = 4) +
    ns(rooms_per_household,      df = 4) +
    ns(population_per_household, df = 4),
  data = train_orig
)

spline_preds <- predict(spline_model, newdata = test_orig)
all_results$spline <- eval_model(test_orig$median_house_value, spline_preds,
                                 "Natural Splines (lm + ns)")
```

Using Cross validation to determine the number of knots

```{r}
spline_vars <- c("median_income", "latitude", "longitude",
                 "housing_median_age", "rooms_per_household",
                 "population_per_household")

best_df <- sapply(spline_vars, function(var) {
  cv_errors <- sapply(3:8, function(df) {
    formula  <- as.formula(paste("median_house_value ~ ns(", var, ", df =", df, ")"))
    fit      <- glm(formula, data = train_orig)
    cv.glm(train_orig, fit, K = 10)$delta[1]
  })
  best <- which.min(cv_errors) + 2  # +2 because we started at df=3
  cat(var, "-> best df:", best, "\n")
  best
})

print(best_df)
```

```{r}
spline_model2 <- lm(
  median_house_value ~
    ns(median_income,            df = best_df["median_income"])            +
    ns(latitude,                 df = best_df["latitude"])                 +
    ns(longitude,                df = best_df["longitude"])                +
    ns(housing_median_age,       df = best_df["housing_median_age"])       +
    ns(rooms_per_household,      df = best_df["rooms_per_household"])      +
    ns(population_per_household, df = best_df["population_per_household"]),
  data = train_orig
)
```

```{r}
spline_preds2 <- predict(spline_model2, newdata = test_orig)
all_results$spline2 <- eval_model(test_orig$median_house_value, spline_preds2,
                                 "Natural Splines Best df (lm + ns)")
```


Checking the variance-bias trade off in the spline fitting
```{r}
bv_spline <- lapply(3:8, function(df) {
  fit        <- lm(median_house_value ~ ns(median_income, df = df), data = train_orig)
  train_rmse <- sqrt(mean((train_orig$median_house_value - predict(fit, train_orig))^2))
  test_rmse  <- sqrt(mean((test_orig$median_house_value  - predict(fit, test_orig))^2))
  data.frame(DF = df, Train_RMSE = train_rmse, Test_RMSE = test_rmse)
})

bv_spline_df <- do.call(rbind, bv_spline)

bv_spline_df %>%
  pivot_longer(cols = c(Train_RMSE, Test_RMSE),
               names_to = "Set", values_to = "RMSE") %>%
  ggplot(aes(x = DF, y = RMSE, colour = Set)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 3:8) +
  labs(title = "Bias-Variance Trade-off: Spline df vs RMSE",
       x = "Degrees of Freedom", y = "RMSE") +
  theme_minimal()
```

The RMSE of the train falls steadily but for the test set levels off. This shows some presence of variance. Since it is not rising, we may not have overfitting challenges.

The degrees of freedom in spline determination help in controlling the bias and variance.


## GNPR — Kernel Regression 

```{r}
library(np)

np_cols <- c("median_income", "latitude", "longitude")

set.seed(42)
sub_idx  <- sample(nrow(train_orig), 2000)
np_train <- train_orig[sub_idx, np_cols]
np_test  <- test_orig[, np_cols]

# Step 1: get bandwidths from training data
bw <- npregbw(
  xdat = np_train,
  ydat = train_orig$median_house_value[sub_idx],
  regtype = "lc",
  bwmethod = "cv.ls",
  tol  = 0.1,
  ftol = 0.1
)

# Step 2: predict using bw object, passing xdat and txdat explicitly
kern_fit <- npreg(
  bws  = bw,
  txdat = np_train,
  tydat = train_orig$median_house_value[sub_idx],
  exdat = np_test
)

kern_preds <- fitted(kern_fit)
all_results$kernel <- eval_model(
  test_orig$median_house_value, kern_preds, "Kernel Regression (np)"
)
```


## GNPR — Local Polynomial Regression (LOESS)

LOESS fits a low-degree polynomial in a *moving window* around each point — the prediction at any location is influenced mainly by nearby observations. It is purely data-driven but struggles in high dimensions, so we restrict it to the three strongest predictors.

```{r loess-model}
loess_model <- loess(
  median_house_value ~ median_income + latitude + longitude,
  data   = train_orig,
  span   = 0.1,    # window width (10 % of data used per local fit)
  degree = 2
)

loess_preds <- predict(loess_model, newdata = test_orig)
loess_preds[is.na(loess_preds)] <- mean(train_orig$median_house_value)
all_results$loess <- eval_model(test_orig$median_house_value, loess_preds,
                                "Local Polynomial Regression (LOESS)")
```

## GNPR — Generalised Additive Model (GAM)

GAM replaces each linear term with a smooth function `s(x)` whose shape is estimated from the data. The tensor product smooth `ti(median_income, latitude)` explicitly models the nonlinear income–location interaction identified in Question 2.

```{r gam-model}
gam_model <- gam(
  median_house_value ~
    s(median_income,            k = 10) +
    s(latitude,                 k = 10) +
    s(longitude,                k = 10) +
    s(housing_median_age,       k = 8)  +
    s(rooms_per_household,      k = 8)  +
    s(bedrooms_per_room,        k = 8)  +
    s(population_per_household, k = 8)  +
    ti(median_income, latitude, k = 5),
  data   = train_orig[, c(MODEL_FEATURES, "median_house_value")],
  method = "REML"
)

gam_preds <- predict(gam_model, newdata = test_orig)
all_results$gam <- eval_model(test_orig$median_house_value, gam_preds,
                              "GAM (mgcv + interaction)")
summary(gam_model)
```

### GAM Smooth Plots

Each curve shows the estimated nonlinear relationship between one predictor and house value (with confidence band). Flat curves mean little effect; curved or wiggly curves mean the relationship is genuinely nonlinear.

```{r gam-plots, fig.height=9}
plot(gam_model, pages = 1, shade = TRUE, seWithMean = TRUE,
     main = "GAM — Estimated Smooth Functions")
```


## GNPR - GAM with Spatial smooth
```{r gam-spatial}
# GAM with 2D spatial smooth over longitude × latitude
# te() models the joint geographic surface — not two separate additive effects
# This directly captures the spatial clustering visible in the EDA map

gam_spatial <- gam(
  median_house_value ~
    s(median_income,            k = 10) +
    s(housing_median_age,       k = 8)  +
    s(rooms_per_household,      k = 8)  +
    s(bedrooms_per_room,        k = 8)  +
    s(population_per_household, k = 8)  +
    te(longitude, latitude,     k = 10) +          # 2D spatial surface
    ti(median_income, latitude, k = 5),             # income × location interaction
  data   = train_orig[, c(MODEL_FEATURES, "median_house_value")],
  method = "REML"
)

gam_spatial_preds <- predict(gam_spatial, newdata = test_orig)
all_results$gam_spatial <- eval_model(
  test_orig$median_house_value, gam_spatial_preds,
  "GAM (spatial te + interaction)"
)

summary(gam_spatial)
```


```{r gam-spatial-surface, fig.height=6}
# Visualise the 2D geographic price surface estimated by the GAM
# This shows the spatial pattern of house values after controlling for
# income, age, and household composition

vis.gam(
  gam_spatial,
  view      = c("longitude", "latitude"),
  plot.type = "contour",
  color     = "terrain",
  n.grid    = 50,
  main      = "GAM Spatial Surface: Geographic Price Structure",
  xlab      = "Longitude", ylab = "Latitude"
)
```
```{r gam-spatial-vs-base}
# Compare spatial GAM against the additive GAM side by side
cat("Additive GAM (s(lat) + s(lon) separately):\n")
print(all_results$gam)

cat("\nSpatial GAM (te(lon, lat) joint surface):\n")
print(all_results$gam_spatial)

```

---

# Final Comparison

## Results Table

All models evaluated on the same held-out 20% test set, same seven features, same dollar-scale response. Sorted by RMSE (lower is better).

```{r results-table}
results_df <- do.call(rbind, all_results) %>%
  arrange(RMSE)
rownames(results_df) <- NULL

# Add model family labels for grouped comparison
results_df$Family <- case_when(
  results_df$Model %in% c("Linear Regression (baseline)",
                           "Polynomial Regression (degree per CV)",
                          "Polynomial (deg-4)",
                          "Polynomial (deg-6)")        ~ "Parametric",
  results_df$Model %in% c("Step Function (income bins)",
                          "Natural Splines (lm + ns)",
                          "Natural Splines Best df (lm + ns)",
                          "Kernel Regression (np)",
                           "Local Polynomial Regression (LOESS)",
                          
                            "GAM (mgcv + interaction)",
                          "GAM (spatial te + interaction)")             ~ "GNPR",
  TRUE                                                              ~ "Machine Learning"
)

knitr::kable(
  results_df[, c("Family", "Model", "RMSE", "MAE", "R2")],
  caption = "Model Performance on Test Set (sorted by RMSE)"
)
```

## RMSE & MAE Bar Chart

```{r results-bar, fig.height=5}
results_df %>%
  pivot_longer(cols = c(RMSE, MAE), names_to = "Metric", values_to = "Value") %>%
  mutate(Model = reorder(Model, -Value)) %>%
  ggplot(aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Model Comparison: RMSE and MAE (lower is better)",
       x = NULL, y = "Error ($)") +
  theme_minimal()
```

## R² Bar Chart

```{r results-r2, fig.height=4}
results_df %>%
  mutate(Model = reorder(Model, R2)) %>%
  ggplot(aes(x = Model, y = R2, fill = Model)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "red") +
  labs(title = "Model Comparison: R² (higher is better — dashed line = 0.80)",
       x = NULL, y = "R²") +
  theme_minimal() +
  theme(legend.position = "none")
```
---

# Conclusions

## Question 1 — Can nonlinear ensembles predict house prices more accurately?

**Yes.** Random Forest and XGBoost consistently outperform linear regression on every metric. Key findings:
- **`median_income`** is the single strongest predictor in both RF and XGBoost importance rankings.
- **`latitude` and `longitude`** rank highly, confirming a strong geographic effect.

## Question 2 - Nonlinear Interactions Between Location and Income

- The **partial dependence interaction plot** shows that the income effect is *steeper* at higher latitudes, a genuine nonlinear interaction that linear models cannot capture.

## Question 3  — Which model wins overall?

| Category        | Model                  | Verdict                                             |
|-----------------|------------------------|-----------------------------------------------------|
| Parametric      | Linear Regression      | Weakest — misses all curvature                      |
| GNPR            | Natural Splines        | Better than OLS; captures smooth nonlinearity       |
| GNPR            | LOESS                  | Good locally but struggles outside training support |
| GNPR            | **GAM**                | Best among GNPR — handles interactions via `ti()`   |
| Machine Learning| Neural Network         | Competitive but needs tuning                        |
| Machine Learning| **XGBoost / RF**       | **Best overall** — lowest RMSE, highest R²          |

**XGBoost (and ensemble of XGBoost and Random Forest) provides the best predictive accuracy.** GAM is the best *interpretable* nonlinear model and is strongly preferred when you need to explain *why* prices are high, rather than just predict that they will be.

#### In terms of bias variance checks?

| Model                                 | Bias   | Variance | Notes                                               |
| :-------------------------------------| :----- | :------- | :---------------------------------------------------|
| **Linear Regression**                 | High   | Low      | Underfits — misses all curvature                    |
| **Polynomial deg-6**                  | Low    | High     | Flexible but risks overfitting (as seen in the CV) |
| **Natural Splines (CV-selected df)**  | Medium | Medium   | Balanced — CV finds sweet spot                      |
| **LOESS (span=0.1)**                  | Low    | High     | Very local fits, unstable at edges                  |
| **GAM**                               | Low    | Medium   | Smooth selection via REML controls variance         |
| **XGBoost / RF**                      | Low    | Low      | Ensemble averaging reduces variance                 |


**XGBoost, an Ensemble of XGBoost and Random Forest and Random Forest itself still perform better as ensemble methods since they average out variances and avoid learnt biases.
